Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2017-08-14 threshold/draft.tex
Typos and other minor miscellaneous edits

Three practical challenges need to be addressed for Algorithm \ref{prune} to be
used in practice:
\begin{enumerate}
\item The algorithm needs a computationally efficient way to compute
a low-rank truncated SVD of potentially large sparse or dense matrices.
\item The truncated SVD method should be able to cheaply restart to
compute additional singular vectors as needed by step 5 of Algorithm
\ref{prune}.
\item Algorithm \ref{prune} as stated assumes a standardized matrix;
it needs to be extended to efficiently apply to general matrices.
\end{enumerate}
%For Algorithm \ref{prune} to be used in practice, three challenges need to be 
%addressed. First, the algorithm requires an efficient method for computing a 
%relatively low-rank truncated SVD of potentially large matrices required by step 
%1. Second, the truncated SVD method should be able to cheaply restart to compute 
%additional singular vectors as required in step 5 of Algorithm \ref{prune}. 
%Third, while the algorithm assumes a standardized matrix, it should work for 
%general matrices. Fortunately, each of these challenges is met with a single 
%algorithm.

The first two challenges are addressed directly by the augmented implicitly
restarted Lanczos bidiagonalization algorithm (IRLBA) of Baglama and Reichel
\cite{irlba}. It efficiently computes truncated singular value decompositions
of either large dense or sparse matrices and the result from an existing rank
$k$ approximation can be efficiently extended to higher-rank approximations by
restarting.  Reference implementations are available for \proglang{R}
\cite{irlbar}, \proglang{Matlab} \cite{irlbam}, and \proglang{Python}
\cite{irlbap}.

@@

A procedure that finds the top $N$ most correlated pairs for a fixed integer
$N$ proceeds iteratively starting with a small initial value of $p$, for
instance $p = 0.999$.  Recall that the maximum number of pairs meeting a
threshold is available by step 4 of Algorithm \ref{prune}.  Then:

@@

The procedure then relies on the ability to cheaply restart IRLBA to expand an
existing truncated SVD.

@@

\begin{Schunk}
\begin{Sinput}
> devtools::install_github("bwlewis/tcor")
\end{Sinput}
\end{Schunk}

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \cite{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95. The choice of projection dimension $p$ affects
the efficiency of the algorithm. For $p=2$ the number of candidate vector pairs
generated in step 4 of Algorithm \ref{prune} is over 3 million. When $p$ is increased
to $10$, the number of candidate vector pairs drops to less than 13,000.
The reference \proglang{R} implementation includes options for
tuning $p$ and relies on the restart ability of the IRLBA to pick up where it left off and
inexpensively extend an existing truncated SVD. The following example uses $p=10$.
\begin{small}
\begin{Schunk}
\begin{Sinput}
threshold = 0.95
library(tcor)
library(doMC)     # local parallel computation
library(biclust)  # for the BicatYeast data
data(EisenYeast)  # in gene by sample orientation
A = t(EisenYeast) # to correlate genes to genes
# Register a four-core multicore parallel backend
registerDoMC(4)

@@

The results computed on the Linux cluster are marked ``tcor (64 core cluster)''
below. The cluster example computed the filtering step 7 of Algorithm
\ref{prune} in parallel across the 64 remote \proglang{R} processes. That approach incurs,
in this example, an additional memory overhead of about 16$\,$GB because each
remote \proglang{R} process requires access to a copy of the data matrix $A$. More thrifty
adaptations of the algorithm are possible and could, for example, share the
memory required by $A$ between worker \proglang{R} processes on each computer.

>>>>

Three practical challenges need to be addressed for Algorithm \ref{prune} to be
used in practice:
\begin{enumerate}
\item The algorithm needs a computationally efficient way to compute
a low-rank truncated SVD of potentially large sparse or dense matrices.
\item The truncated SVD method should be able to cheaply restart to
compute additional singular vectors as needed by step 5 of Algorithm
\ref{prune}.
\item Algorithm \ref{prune} as stated assumes a standardized matrix;
it needs to be extended to general matrices.
\end{enumerate}
%For Algorithm \ref{prune} to be used in practice, three challenges need to be 
%addressed. First, the algorithm requires an efficient method for computing a 
%relatively low-rank truncated SVD of potentially large matrices required by step 
%1. Second, the truncated SVD method should be able to cheaply restart to compute 
%additional singular vectors as required in step 5 of Algorithm \ref{prune}. 
%Third, while the algorithm assumes a standardized matrix, it should work for 
%general matrices. Fortunately, each of these challenges is met with a single 
%algorithm.

The first two challenges are addressed directly by the augmented implicitly
restarted Lanczos bidiagonalization algorithm (IRLBA) of Baglama and Reichel
\cite{irlba}. It efficiently computes truncated singular value decompositions
of either large dense or sparse matrices. And, the result from an existing rank
$k$ approximation can be efficiently extended to higher-rank approximations by
restarting.  Reference implementations are available for \proglang{R}
\cite{irlbar}, \proglang{Matlab} \cite{irlbam}, and \proglang{Python}
\cite{irlbap}.

@@

A procedure that finds the top $N$ most correlated pairs for a fixed integer
$N$ proceeds iteratively starting with a small initial value of $p$, for
instance $p = 0.999$.  The maximum number of pairs meeting a
threshold is available by step 4 of Algorithm \ref{prune}.  Then:

@@

The procedure relies on the ability to cheaply restart IRLBA to expand an
existing truncated SVD.

@@

\begin{Schunk}
\begin{Sinput}
devtools::install_github("bwlewis/tcor")
\end{Sinput}
\end{Schunk}

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \cite{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95. The choice of projection dimension $p$ affects
the efficiency of the algorithm. For $p=2$ the number of candidate vector pairs
generated in step 4 of Algorithm \ref{prune} is over 3 million. When $p$ is increased
to $10$, the number of candidate vector pairs drops to less than 13,000.
The reference \proglang{R} implementation includes options for
tuning $p$ and relies on the restart ability of the IRLBA to pick up where it left off and
inexpensively extend an existing truncated SVD. The following example uses $p=10$.
\begin{small}
\begin{Schunk}
\begin{Sinput}
threshold = 0.95
library(tcor)
library(doMC)     # local parallel computation
library(biclust)  # for the BicatYeast data
data(EisenYeast)
A = t(EisenYeast) # (to correlate genes to genes)
registerDoMC(4)   # register a four-core multicore parallel backend

@@

The results computed on the Linux cluster are marked ``tcor (64 core cluster)''
below. The cluster example computed the filtering step 7 of Algorithm
\ref{prune} in parallel across the 64 remote \proglang{R} processes. That
approach incurs, in this example, an additional memory overhead because each
remote \proglang{R} process requires access to a copy of the data matrix $A$.
More thrifty adaptations of the algorithm are possible and could, for example,
share the memory required by $A$ between worker \proglang{R} processes on each
computer.
